---
title: "Lab 09: Algorithmic Bias"
author: "Tiffani J. Hill"
date: "03.04.2026"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(janitor)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r}
compas <- read_csv("data/compas-scores-2-years.csv") %>% 
  clean_names() %>% 
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )
glimpse(compas)
```
###Exercise 1
There are 7214 rows, which represent individual observations. There are 53 columns. 
```{r}
nrow(compas)
ncol(compas)
colnames(compas)

```

## Exercise 2
The number of defendants is equivalent to the number of rows. Each entry has a unique id number.

###Exercise 3
```{r risk-score}

compas %>% 
  ggplot(aes(x = decile_score)) +
  geom_histogram()


```
###Exercise 4
```{r visual-by-demographics}
compas %>% 
  ggplot(aes(x = decile_score)) +
  geom_histogram() +
  facet_wrap(~sex)
  

compas %>% 
  ggplot(aes(x = decile_score)) +
  geom_histogram() +
  facet_wrap(~race)

compas %>% 
  ggplot(aes(x = decile_score)) +
  geom_histogram() +
  facet_wrap(~age_cat)
```
```{r challenge-visual}
compas %>% 
  ggplot(aes(x = decile_score, color = race))+
  geom_bar() +
  facet_wrap(~age_cat)
```
```{r recidivisism}
compas %>% 
  ggplot(aes(x = two_year_recid, y = decile_score)) +
geom_count()
```
Not exactly. 

```{r compas-accuracy}

compas_accuracy <- lm(compas$two_year_recid ~ compas$decile_score)
summary(compas_accuracy)

cor(compas$decile_score, compas$two_year_recid)
```