---
title: "Lab 09: Algorithmic Bias"
author: "Tiffani J. Hill"
date: "03.04.2026"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(janitor)
library(dplyr)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r}
compas <- read_csv("data/compas-scores-2-years.csv") %>% 
  clean_names() %>% 
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )
glimpse(compas)
```
###Exercise 1
There are 7214 rows, which represent individual observations. There are 53 columns. 
```{r}
nrow(compas)
ncol(compas)
colnames(compas)

```

## Exercise 2
The number of defendants is equivalent to the number of rows. Each entry has a unique id number.

###Exercise 3
```{r risk-score}

compas %>% 
  ggplot(aes(x = decile_score)) +
  geom_histogram()


```
###Exercise 4
```{r visual-by-demographics}
compas %>% 
  ggplot(aes(x = decile_score)) +
  geom_histogram() +
  facet_wrap(~sex)
  

compas %>% 
  ggplot(aes(x = decile_score)) +
  geom_histogram() +
  facet_wrap(~race)

compas %>% 
  ggplot(aes(x = decile_score)) +
  geom_histogram() +
  facet_wrap(~age_cat)
```
```{r challenge-visual}
compas %>% 
  ggplot(aes(x = decile_score, color = race))+
  geom_bar() +
  facet_wrap(~age_cat)
```
```{r recidivisism}
compas %>% 
  ggplot(aes(x = two_year_recid, y = decile_score)) +
geom_count()
```
Not exactly. 
The COMPAS algorithim is correct 55% of the time.
```{r compas-accuracy}
compas <- compas %>% 
  mutate(
    compas_classification = case_when(decile_score >= 7 & two_year_recid == 1 ~ "TP",
   decile_score <= 4 & two_year_recid == 0 ~"TN",
   decile_score >= 7 & two_year_recid == 0 ~ "FP",
   decile_score <= 4 & two_year_recid == 1 ~ "FN")  )

compas_summary <- compas %>%
  filter(!is.na(compas_classification)) %>%
  count(compas_classification)

compas %>%
  count(compas_classification) %>%
  mutate(percent = n / sum(n) * 100)

ggplot(compas_summary, aes(x = compas_classification, y = n)) +
  geom_col() +
  labs(
    title = "COMPAS Prediction Outcomes",
    x = "Classification",
    y = "Number of Cases"
  ) +
  theme_minimal()
```
##Part 3
More white defendants are deemed low-risk than Black defendants. For white people, there seems to be a positive skew. 
```{r investigating-disparities}
filterbw <-
compas %>% 
  filter(race == "African-American"| race =="Caucasian") 

filterbw %>% 
  ggplot(aes(x = decile_score)) +
  geom_histogram() +
  facet_grid(~ race)
```

There's clearly a disparity between Black defendants and white defendants. About 39% of Black defendants are considered high risk, while 17% of white defendants are considered high risk.

```{r disparities-percent}
filterbw <- filterbw %>% 
  mutate(high_risk = decile_score >= 7)

filterbw %>% 
  group_by(race) %>% 
  summarise(
    total_defendants = n(),
    high_risk_count = sum(high_risk),
    percent_high_risk = mean(high_risk) * 100)
  


```


```{r disparities-accuracy}
filterbw <- filterbw %>% 
  mutate(accuracy = compas_classification == "TP" | compas_classification == "TN")

filterbw %>% 
  group_by(race) %>% 
  summarise(
    total_defendants = n(),
    accurate_predictions = sum(accuracy, na.rm = TRUE),
    percent_accurate = mean(accuracy, na.rm = TRUE) * 100)
```
  )


```{r accuracy-by-group}
non_recidivists <- compas %>%
  filter(two_year_recid == 0)

non_recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  summarise(
    total_nonrecid = n(),
    false_positives = sum(decile_score >= 7, na.rm = TRUE),
    false_positive_rate = mean(decile_score >= 7, na.rm = TRUE) * 100
  )

recidivists <- compas %>%
  filter(two_year_recid == 1)

recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  summarise(
    total_recid = n(),
    false_negatives = sum(decile_score <= 4, na.rm = TRUE),
    false_negative_rate = mean(decile_score <= 4, na.rm = TRUE) * 100
  )
```
There is a much larger false negative rate for Caucasian defendants than Black defendants. Conversely, there is a much larger false positive rate for Black defendants than white defendants.
```{r disparity-visualization}

fpr <- compas %>%
  filter(two_year_recid == 0,
         race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  summarise(rate = mean(decile_score >= 7, na.rm = TRUE) * 100) %>%
  mutate(metric = "False Positive Rate")

fnr <- compas %>%
  filter(two_year_recid == 1,
         race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  summarise(rate = mean(decile_score <= 4, na.rm = TRUE) * 100) %>%
  mutate(metric = "False Negative Rate")

rates <- bind_rows(fpr, fnr)

ggplot(rates, aes(x = race, y = rate, fill = metric)) +
  geom_col(position = "dodge") +
  labs(
    title = "COMPAS Prediction Errors by Race",
    x = "Race",
    y = "Percentage",
    fill = "Metric"
  ) +
  theme_minimal()
```
```