---
title: "Lab 09: Algorithmic Bias"
author: "Tiffani J. Hill"
date: "03.04.2026"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(janitor)
library(dplyr)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r}
compas <- read_csv("data/compas-scores-2-years.csv") %>% 
  clean_names() %>% 
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )
glimpse(compas)
```
###Exercise 1
There are 7214 rows, which represent individual observations. There are 53 columns. 
```{r}
nrow(compas)
ncol(compas)
colnames(compas)

```

## Exercise 2
The number of defendants is equivalent to the number of rows. Each entry has a unique id number.

###Exercise 3
```{r risk-score}

compas %>% 
  ggplot(aes(x = decile_score)) +
  geom_histogram()


```
###Exercise 4
```{r visual-by-demographics}
compas %>% 
  ggplot(aes(x = decile_score)) +
  geom_histogram() +
  facet_wrap(~sex)
  

compas %>% 
  ggplot(aes(x = decile_score)) +
  geom_histogram() +
  facet_wrap(~race)

compas %>% 
  ggplot(aes(x = decile_score)) +
  geom_histogram() +
  facet_wrap(~age_cat)
```
```{r challenge-visual}
compas %>% 
  ggplot(aes(x = decile_score, color = race))+
  geom_bar() +
  facet_wrap(~age_cat)
```
```{r recidivisism}
compas %>% 
  ggplot(aes(x = two_year_recid, y = decile_score)) +
geom_count()
```
Not exactly. 
The COMPAS algorithim is correct 55% of the time.
```{r compas-accuracy}
compas <- compas %>% 
  mutate(
    compas_classification = case_when(decile_score >= 7 & two_year_recid == 1 ~ "TP",
   decile_score <= 4 & two_year_recid == 0 ~"TN",
   decile_score >= 7 & two_year_recid == 0 ~ "FP",
   decile_score <= 4 & two_year_recid == 1 ~ "FN")  )

compas_summary <- compas %>%
  filter(!is.na(compas_classification)) %>%
  count(compas_classification)

compas %>%
  count(compas_classification) %>%
  mutate(percent = n / sum(n) * 100)

ggplot(compas_summary, aes(x = compas_classification, y = n)) +
  geom_col() +
  labs(
    title = "COMPAS Prediction Outcomes",
    x = "Classification",
    y = "Number of Cases"
  ) +
  theme_minimal()
```
##Part 3
More white defendants are deemed low-risk than Black defendants. For white people, there seems to be a positive skew. 
```{r investigating-disparities}
filterbw <-
compas %>% 
  filter(race == "African-American"| race =="Caucasian") 

filterbw %>% 
  ggplot(aes(x = decile_score)) +
  geom_histogram() +
  facet_grid(~ race)
```

There's clearly a disparity between Black defendants and white defendants. About 39% of Black defendants are considered high risk, while 17% of white defendants are considered high risk.

```{r disparities-percent}
filterbw <- filterbw %>% 
  mutate(high_risk = decile_score >= 7)

filterbw %>% 
  group_by(race) %>% 
  summarise(
    total_defendants = n(),
    high_risk_count = sum(high_risk),
    percent_high_risk = mean(high_risk) * 100)
  


```


```{r disparities-accuracy}
filterbw <- filterbw %>% 
  mutate(accuracy = compas_classification == "TP" | compas_classification == "TN")

filterbw %>% 
  group_by(race) %>% 
  summarise(
    total_defendants = n(),
    accurate_predictions = sum(accuracy, na.rm = TRUE),
    percent_accurate = mean(accuracy, na.rm = TRUE) * 100)
```
  )


```{r accuracy-by-group}
non_recidivists <- compas %>%
  filter(two_year_recid == 0)

non_recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  summarise(
    total_nonrecid = n(),
    false_positives = sum(decile_score >= 7, na.rm = TRUE),
    false_positive_rate = mean(decile_score >= 7, na.rm = TRUE) * 100
  )

recidivists <- compas %>%
  filter(two_year_recid == 1)

recidivists %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  summarise(
    total_recid = n(),
    false_negatives = sum(decile_score <= 4, na.rm = TRUE),
    false_negative_rate = mean(decile_score <= 4, na.rm = TRUE) * 100
  )
```
There is a much larger false negative rate for Caucasian defendants than Black defendants. Conversely, there is a much larger false positive rate for Black defendants than white defendants.
```{r disparity-visualization}

fpr <- compas %>%
  filter(two_year_recid == 0,
         race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  summarise(rate = mean(decile_score >= 7, na.rm = TRUE) * 100) %>%
  mutate(metric = "False Positive Rate")

fnr <- compas %>%
  filter(two_year_recid == 1,
         race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  summarise(rate = mean(decile_score <= 4, na.rm = TRUE) * 100) %>%
  mutate(metric = "False Negative Rate")

rates <- bind_rows(fpr, fnr)

ggplot(rates, aes(x = race, y = rate, fill = metric)) +
  geom_col(position = "dodge") +
  labs(
    title = "COMPAS Prediction Errors by Race",
    x = "Race",
    y = "Percentage",
    fill = "Metric"
  ) +
  theme_minimal()
```
Unfortunately, this is not the best graph, but after looking at it for a second, one can see that prior convictions seem to have a stronger correlation with decile score for Black defendants than any other racial group. For Black and White defendants, this is where prior convictions even seem to matter at all, but for White defendants it is more likely to get a lower decile score than Black defendants.
```{r bias-visual}
compas %>% 
  ggplot(aes(x = priors_count, y = decile_score,
             color = race)) +
  geom_count(alpha = 1) +
  facet_wrap(~ race)
```
#calculate recidivism rates by score for each race

```{r recidivism-rates}
compas %>% 
  group_by(race) %>% 
  summarise(
    total_people = n(),
    total_recid = sum(two_year_recid, na.rm = TRUE),
    percent_recid = mean(two_year_recid, na.rm = TRUE) * 100
  )

```

If the algorithm was fair, the lines would mostly overlap. However, this is not the case. For Caucasion and African American defendants, the lines mostly overlap but for all other races they seem to vary quite drastically. Thus, the algorithm doesn't treat all races the same.
```{r plot-recid}
library(dplyr)

calibration <- compas %>%
  group_by(race, decile_score) %>%
  summarise(
    n = n(),
    recid_rate = mean(two_year_recid, na.rm = TRUE)
  ) %>%
  ungroup()

library(ggplot2)

ggplot(calibration, aes(x = decile_score, y = recid_rate, color = race)) +
  geom_line() +
  geom_point() +
  labs(
    x = "COMPAS Risk Score (Decile)",
    y = "Observed Recidivism Rate",
    title = "Calibration of COMPAS Risk Scores by Race"
  ) +
  theme_minimal()
```
To create a fairer risk assessment algorithm, I would remove/modify variables that indirectly encode for race that would create bias. Additionally, it would be essential to conduct regular bias audits, something that researchers like us could do, to identify and address these disparities if they do arise. 

Fairness goals can conflict with each other when attempting to create a well calibrated system but also creating equal error rates across racial groups. A well calibrated algorithm would mean the same risk score has similar probabilities of recidivism across demographic groups. However, this could create unequal error rates in some racial groups. To equalize this, this could create an algorithm that is no longer perfectly calibrated. Additionally, increasing predictive accuracy may reduce fairness. 

It is imperative for algorithmic risk assessments to only be used as one factor among many when making judicial decisions. Judges and legal professionals should also receive training on how to interpret algorithmic predictions and understand their limitations. 